The effect of teachers reassigning students to new Cognitive Tutor sections

### 2020-06-01 paper.Rnw
small changes to text

\maketitle
\begin{abstract}
The design of the Cognitive Tutor Algebra I (CTA1) intelligent
tutoring system assumes that students work through sections of
material following a pre-specified order, and only move on from one
section to the next after mastering the first section’s
skills. However, the software gives teachers the ability to override
that structure, by reassigning students to different sections of the
curriculum. Which students get reassigned? Does reassignment hurt
student learning? Does it help? This paper used data from the
treatment arm of a large effectiveness study of the CTA1 curriculum to
estimate the effects of reassignment on students’ scores on an Algebra
I posttest. Since reassignment is not randomized, we used a multilevel
propensity score matching design, along with assessments of
sensitivity to bias from unmeasured confounding, to estimate the
effects of reassignment. We found that reassignment reduces posttest
scores by roughly 0.2 standard deviations—about the same as the overall CTA1 treatment effect—that unmeasured confounding is unlikely to completely explain this observed effect, but that the effect of reassignment may vary widely between classrooms.
\end{abstract}

@@

\section{Introduction}
Two closely related pillars of intelligent tutoring systems are
sequencing and mastery learning.
It has long been obvious that the sequence in which students learn
different topics is an important component of a curriculum, due to
prerequisites---for instance, students must master arithmetic in order
to learn how to solve algebraic equations.
A related example is scaffolding, in which
learners gradually achieve independence over a sequence of problems;
scaffolding ``consists essentially of the adult 'controling' those
elements of the task that are initially beyond the learner's capacity,
thus permitting him to concentrate upon and complete only those
elements that are within his range of competence''
\cite{wood1976role}.
However, measuring the effects of sequencing \cite{ritter2007order}
\cite{doroudi2016sequence} and determining prerequisites or optimal
sequences \cite{scheines2014discovering} \cite{vuong2011method} \cite{koedinger2002toward}
remains an active area of research.

@@

The following section gives background on the effectiveness trial
and describe the data we will use for the study.
Section \ref{sec:psm} describes propensity score matching, the
method we employ. Section \ref{sec:propensityScores} describes the
propensity score models, which in turn describe characteristics of
students who are reassigned. Section \ref{sec:match} describes the
matching algorithm and covariate balance. Section \ref{sec:effects}
gives our main results on the effects of reassignment, including
sensitivity analysis to confounding from unmeasured covariates and
between-classroom effect heterogeneity. Section \ref{sec:discussion} concludes.

@@

Unfortunately, log data was not available for every student in the
treatment arm of the study, primarily for two reasons: some students
in CTA1 schools nevertheless did not use the tutor, and some students
used the tutor but their log data was irretrievable or could not be
reliably linked to posttest scores and covariates.
This study omitted schools in which data was missing for over 20\% of
students in either year, leaving \Sexpr{pn(nschool)} schools.
Among the students at these schools, we omitted 164 had no log data, and 242
either only worked---but did not complete---one section or had no
section completion data for some other reason.
A total of \Sexpr{pn(nstud)} students in \Sexpr{pn(nclass)} classrooms
remained in the analysis sample, roughly \Sexpr{percentKept}\% of the
full treatment group.

@@

\citeN{hhh} suggest a method of estimating the sensitivity of a
regression to an omitted confounder based on benchmarking from
observed confounders.
Roughly speaking, the idea is to widen the confidence interval from an
ostensibly causal linear model to account for the possibility of a
hypothetical unmeasured confounder, $U$, that predicts reassignment and posttests to the same extent as one of the observed covariates.
These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.

In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
Capturing these two requirements, the method of \citeN{hhh}
is based on two sensitivity parameters: first, $T_Z$ encodes the
extent to which $U$ predicts $Z$, after accounting for observed
covariates $\bm{x}$.
Formally, $T_Z$ is the t-statistic on the $U$ coefficient from an ordinary least squares
regression of $Z$ on $U$ and $\bm{X}$.
The second parameter is $\rho^2$, the squared partial correlation
between posttest scores and $U$, conditional on $\bm{x}$.
Of course, since $U$ is unobserved, neither $T_Z$ nor $\rho^2$ is
known; \citeN{hhh} suggest benchmarking them using observed
covariates.
That is, imagine each observed covariate, in turn, were unobserved,
and calculate its $T_Z$ and $\rho^2$ given the rest of the observed
covariates.

@@

If reassignment is used differently from classroom to classroom, it
stands to reason that it might have different effects in different
classrooms, as well.
To test that assumption, we fit a multilevel model with random effects
for reassignment, varying by classroom.
The model had the same fixed effects as model underlying the
``Match+Regression'' results described above, as well as random
intercepts for classroom and random slopes for reassignment, varying
by classroom.
Formally, the model is:
\begin{equation}\label{eq:heteroReg}
\begin{split}
y_i&=\beta_{0,m[i]}+\beta_1Z_i+\\
&\beta_2SpEd_i+\beta_3frlMIS_i+\beta_4Black_i+\beta_5Hisp_i+\\
&ns^5(pretest_i,\bm{\alpha})+\gamma_{j[i]}Z_i+\epsilon^{Cls}_{j[i]}+\epsilon^{Ind}_i
\end{split}
\end{equation}
where $\beta_{0,m[i]}$ is a fixed intercept for each match, $ns^5(pretest_i,\bm{\alpha})$ is a natural spline for pretest, with five degrees of freedom and coefficient vector $\alpha$, and $\gamma_{j[i]}$, $\epsilon^{Cls}_{j[i]}$, and $\epsilon^{Ind}_i$ are random effects, modeled as normal with mean zero and standard deviation estimated from the data.
Symbols $\alpha$, $\beta$, $\gamma$, and $\epsilon$ do not represent the same quantities as in equation \eqref{eq:propmod1}. 
$\gamma_{j[i]}$ is the random slope for reassignment, varying by classroom; the effect of reassignment in classroom $j$ is estimated as $\hat{\beta}_1+\hat{\gamma}_j$.
That is, $\beta_1$ represents the effect of reassignment, averaged over all classrooms, and $\gamma_j$ represents the difference between classroom $j$'s effect and the average. 
While precisely estimating the effect of reassignment in any particular
classroom is beyond the scope of our data, this model allows us to estimate the variance across those effects, as the variance of $\gamma_j$s.

@@

This variation could be due to a number of factors, including
differences in the composition of classroomsand in when and how reassignment is used.
We considered two simple hypotheses about classroom-level predictors of heterogeneous treatment effects.
The first hypothesis was that variance in students' pretest scores within a classroom predicts the effect of reassignment in that classroom.
The idea is that some teachers may use reassignment as a tool to address varying student ability---for instance, they may reassign lagging students to help them keep up with their classmates.
Classrooms with higher variance in pretest scores afford more opportunities for teachers to use this reassignment strategy. 
If the strategy is widely used, and either particularly effective or ineffective at boosting students' posttest scores, there will be a correlation between classroom-level variance in pretest scores and the effect of reassignment.

Our second hypothesis was that the proportion of students in a classroom who have been reassigned may predict classroom-level effects.
The idea here is that in classrooms with low a low proportion of students reassigned, teachers use reassignment in a more targeted fashion, so it may be more beneficial.

>>>>

\maketitle
\begin{abstract}
The design of the Cognitive Tutor Algebra I (CTA1) intelligent
tutoring system assumes that students work through sections of
material following a pre-specified order, and only move on from one
section to the next after mastering the first section’s
skills. However, the software gives teachers the flexibility to override
that structure, by reassigning students to different sections of the
curriculum. Which students get reassigned? Does reassignment hurt
student learning? Does it help? This paper used data from the
treatment arm of a large effectiveness study of the CTA1 curriculum to
estimate the effects of reassignment on students’ scores on an Algebra
I posttest. Since reassignment is not randomized, we used a multilevel
propensity score matching design, along with assessments of
sensitivity to bias from unmeasured confounding, to estimate the
effects of reassignment. We found that reassignment reduces posttest
scores by roughly 0.2 standard deviations—--about the same as the overall CTA1 treatment effect---that unmeasured confounding is unlikely to completely explain this observed effect, and that the effect of reassignment may vary widely between classrooms.
\end{abstract}

@@

\section{Introduction}
Two closely related pillars of intelligent tutoring systems are
sequencing and mastery learning.
It has long been obvious that the sequence in which students learn
different topics is an important component of a curriculum, due to
prerequisites---for instance, students must master arithmetic in order
to learn how to solve algebraic equations.
A related example is scaffolding, in which
learners gradually achieve independence over a sequence of problems;
scaffolding ``consists essentially of the adult `controling' those
elements of the task that are initially beyond the learner's capacity,
thus permitting him to concentrate upon and complete only those
elements that are within his range of competence''
\cite{wood1976role}.
However, measuring the effects of sequencing \cite{ritter2007order}
\cite{doroudi2016sequence} and determining prerequisites or optimal
sequences \cite{scheines2014discovering} \cite{vuong2011method} \cite{koedinger2002toward}
remains an active area of research.

@@

The following section gives background on the effectiveness trial
and describes the data we will use for the study.
Section \ref{sec:psm} describes propensity score matching, the
method we employ. Section \ref{sec:propensityScores} describes the
propensity score models, which in turn describe characteristics of
students who are reassigned. Section \ref{sec:match} describes the
matching algorithm and covariate balance. Section \ref{sec:effects}
gives our main results on the effects of reassignment, including
sensitivity analysis to confounding from unmeasured covariates and
between-classroom effect heterogeneity. Section \ref{sec:discussion} concludes.

@@

Unfortunately, log data was not available for every student in the
treatment arm of the study, primarily for two reasons: some students
in CTA1 schools nevertheless did not use the tutor, and some students
used the tutor but their log data was irretrievable or could not be
reliably linked to posttest scores and covariates.
This study omitted schools in which data was missing for over 20\% of
students in either year, leaving \Sexpr{pn(nschool)} schools.
Among the students at these schools, we omitted 164 who had no log data, and 242
who worked---but did not complete---only one section or who had no
section completion data for some other reason.
A total of \Sexpr{pn(nstud)} students in \Sexpr{pn(nclass)} classrooms
remained in the analysis sample, roughly \Sexpr{percentKept}\% of the
full treatment group.

@@

\cite{hhh} suggest a method of estimating the sensitivity of a
regression to an omitted confounder based on benchmarking from
observed confounders.
Roughly speaking, the idea is to widen the confidence interval from an
ostensibly causal linear model to account for the possibility of a
hypothetical unmeasured confounder, $U$, that predicts reassignment and posttests to the same extent as one of the observed covariates.
These ``sensitivity intervals'' account for uncertainty from two sources: random error, and systematic error due to the omission of a confounder.

In order to confound the causal relationship between reassignment and posttests, a confounder would have to predict both.
Capturing these two requirements, the method of \cite{hhh}
is based on two sensitivity parameters: first, $T_Z$ encodes the
extent to which $U$ predicts $Z$, after accounting for observed
covariates $\bm{x}$.
Formally, $T_Z$ is the t-statistic on the $U$ coefficient from an ordinary least squares
regression of $Z$ on $U$ and $\bm{X}$.
The second parameter is $\rho^2$, the squared partial correlation
between posttest scores and $U$, conditional on $\bm{x}$.
Of course, since $U$ is unobserved, neither $T_Z$ nor $\rho^2$ is
known; \cite{hhh} suggest benchmarking them using observed
covariates.
That is, imagine each observed covariate, in turn, were unobserved,
and calculate its $T_Z$ and $\rho^2$ given the rest of the observed
covariates.

@@

If reassignment is used differently from classroom to classroom, it
stands to reason that it might have different effects in different
classrooms, as well.
To test that assumption, we fit a multilevel model with random effects
for reassignment, varying by classroom.
The model had the same fixed effects as model underlying the
``Match+Regression'' results described above, as well as random
intercepts for classroom and random slopes for reassignment, varying
by classroom.
Formally, the model is:
\begin{equation}\label{eq:heteroReg}
\begin{split}
Post&test_i=\beta_{0,m[i]}+\beta_1Z_i+\beta_2SpEd_i+\\
&\beta_3frlMIS_i+\beta_4Black_i+\beta_5Hisp_i+\\
&ns^5(pretest_i,\bm{\alpha})+\gamma_{j[i]}Z_i+\epsilon^{Cls}_{j[i]}+\epsilon^{Ind}_i
\end{split}
\end{equation}
where $\beta_{0,m[i]}$ is a fixed intercept for each match, $ns^5(pretest_i,\bm{\alpha})$ is a natural spline for pretest, with five degrees of freedom and coefficient vector $\alpha$, and $\gamma_{j[i]}$, $\epsilon^{Cls}_{j[i]}$, and $\epsilon^{Ind}_i$ are random effects, modeled as normal with mean zero and standard deviation estimated from the data.
Symbols $\alpha$, $\beta$, $\gamma$, and $\epsilon$ do not represent the same quantities as in equation \eqref{eq:propmod1}.
$\gamma_{j[i]}$ is the random slope for reassignment, varying by classroom; the effect of reassignment in classroom $j$ is estimated as $\hat{\beta}_1+\hat{\gamma}_j$.
That is, $\beta_1$ represents the effect of reassignment, averaged over all classrooms, and $\gamma_j$ represents the difference between classroom $j$'s effect and the average.
While precisely estimating the effect of reassignment in any particular
classroom is beyond the scope of our data, this model allows us to estimate the variance across those effects, as the variance of $\gamma_j$s.

@@

This variation could be due to a number of factors, including
differences in the composition of classrooms and in when or how reassignment is used.
We considered two simple hypotheses about classroom-level predictors of heterogeneous treatment effects.
The first hypothesis was that variance in students' pretest scores within a classroom predicts the effect of reassignment in that classroom.
The idea is that some teachers may use reassignment as a tool to address varying student ability---for instance, they may reassign lagging students to help them keep up with their classmates.
Classrooms with higher variance in pretest scores afford more opportunities for teachers to use this reassignment strategy.
If the strategy is widely used, and either particularly effective or ineffective at boosting students' posttest scores, there will be a correlation between classroom-level variance in pretest scores and the effect of reassignment.

Our second hypothesis was that the proportion of students in a classroom who have been reassigned may predict classroom-level effects.
The idea here is that in classrooms with a low proportion of students reassigned, teachers use reassignment in a more targeted fashion, so it may be more beneficial.
